---
title: "Multilevel/Mixed Effects Model Packages in R"
output:
    beamer_presentation:
        slide_level: 2
        keep_tex: true
---

```{r, setup, echo=FALSE, warning = FALSE, message = FALSE, results = 'hide'}
library(brms)
library(lme4)
library(nlme)
library(ggplot2)
library(BoutrosLab.plotting.general)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = 'center', out.width = '90%', fig.pos = 'H');

theme_set(theme_classic())
```

## Package choices
  - `lme4`
    - "Fit linear and generalized linear mixed-effects models."
  - `nlme`
    - Linear and Nonlinear Mixed Effects Models
    - Can specify covariance structure for the residuals
      - Useful for complex longitudinal modeling
  - `gamm4`
    - "Generalized Additive Mixed Models using 'mgcv' and 'lme4'"
  - `brms`
    - "Fit Bayesian generalized (non-)linear multivariate multilevel models using 'Stan' for full Bayesian inference." 
    
```{r, eval = F}
# This model has fixed effects only for the interaction, intercept included per default
fit1.lmer <- lmer(weight ~ 1 + Time + (1 | Rat), data = BodyWeight)
BodyWeight$mod1.predict <- predict(fit1.lmer)
BodyWeight$mod1.marginal <- predict(fit1.lmer, re.form = NA)

ggplot(BodyWeight) +
  geom_line(aes(x = Time, y = weight, group = Rat, colour = Diet), alpha = 0.5) +
  geom_line(aes(x = Time, y = mod1.predict, group = Rat))

plot(residuals(fit1.lmer))
acf(residuals(fit1.lmer), )
pacf(residuals(fit1.lmer))

fit2.lme <- lme(
  fixed = weight ~ 1 + Time,
  random = ~ 1 | Rat,
  data = BodyWeight,
  method = "REML")
plot(ACF(fit2.lme, resType = "normalized"))

fit3.lme <- lme(
  fixed = weight ~ 1 + Time,
  random = ~ 1 | Rat,
  correlation = corAR1(form = ~ 1 | Rat),
  data = BodyWeight,
  method = "REML"
  )
plot(ACF(fit2.lme, resType = "normalized"))

anova(fit2.lme, fit3.lme)

BodyWeight$fit3.predict <- predict(fit3.lme, level = 0:1)
ggplot(BodyWeight) +
  geom_line(aes(x = Time, y = weight, group = Rat, colour = Diet), alpha = 0.5) +
  geom_line(aes(x = Time, y = fit3.predict, group = Rat))
```

## sleepstudy
```{r, echo = T}
sleepstudy.plot <- create.scatterplot(
  Reaction ~ Days | Subject,
  data = sleepstudy,
  type = c('p', 'r'),
  layout = c(6, 3),
  xlab.label = "Days of sleep deprivation",
  ylab.label = "Average reaction time (ms)",
  ylab.cex = 1.2,
  xlab.cex = 1.2,
  xaxis.cex = 1,
  yaxis.cex = 1,
  lwd = 2.5
  )
```
## sleepstudy
```{r}
# key = list(
#     space = "right",
#    lines=list(col=default.colours(3), lwd=4),
#         text=list(c("Patient Int Pred","Average", "Patient Int+Slope Pred"))
#  )
sleepstudy.plot
```

## Sleep Study - Model 1
  - Fixed effect for time (Days)
  - Per subject intercept
```{r, echo = T}
# Fit the model using ML
fm1 <- lmer(
  formula = Reaction ~ 1 + Days + (1 | Subject),
  data = sleepstudy,
  REML = FALSE)
# Predict with subject-specific intercepts
sleepstudy$fm1.predict <- predict(fm1)
# Predict 
sleepstudy$fm1.predict.marginal <- predict(
  fm1,
  re.form = NA)
```

## Subject Specific Prediction
```{r}
sleepstudy.plot + create.scatterplot(
  fm1.predict ~ Days | Subject,
  data = sleepstudy,
  type = c('l'),
  col = default.colours(3)[1],
  lwd = 2.5,
  layout = c(6, 3)
) + create.scatterplot(
  fm1.predict.marginal ~ Days | Subject,
  data = sleepstudy,
  type = c('l'),
  lwd = 2.5,
  col = default.colours(3)[2],
  layout = c(6, 3)
)
```
  
## REML vs ML
  - Two main ways of fitting a mixed-effect model: REML (Restricted Maximum-likelihood) and ML 
    - REML has biased estimates for fixed effects when study is unbalanced
    - ML has biased estimates for variance components
  - In `lme4` use `REML = TRUE` or `REML = FALSE`
  - In `nlme` use `method = "ML"` or `method = "REML"`
  - Use ML if comparing fixed effects or REML if comparing variance components
  - Need ML if using likelihood ratio tests (or AIC/BIC)

##  Sleep Study - Model 1 REML
```{r, echo = T}
fm1.REML <- lmer(
  formula = Reaction ~ 1 + Days + (1 | Subject),
  data = sleepstudy,
  REML = TRUE)
```

## Sleep Study - Model 1 REML
```{r}
fm1.REML
```

## lme4 syntax
  - `(1 | Subject)`
    - subject specific intercept
  - `(1 + Days | Subject)`
    - subject specific intercept and slope with Time and Subject *correlated*
  - `(1 | Subject) + (0 + Days | Subject)`
    - subject specific intercept and slope with Time and Subject *independent*
    - Don't forget the `0 + Days` or `(Days | Subject)` implicitly will fit `(1 + Days | Subject)`
  
## Sleep Study - Model 2 (Subject slopes)
```{r, echo = T}
fm2 <- lmer(
  formula = Reaction ~ 1 + Days + (1 + Days | Subject),
  data = sleepstudy,
  REML = FALSE)
sleepstudy$fm2.predict <- predict(fm2)
```

## Sleep Study - Model 2 (Subject slopes)
```{r}
sleepstudy.plot + create.scatterplot(
  fm1.predict ~ Days | Subject,
  data = sleepstudy,
  type = c('l'),
  col = default.colours(3)[1],
  lwd = 2.5,
  layout = c(6, 3)
) + create.scatterplot(
  fm1.predict.marginal ~ Days | Subject,
  data = sleepstudy,
  type = c('l'),
  lwd = 2.5,
  col = default.colours(3)[2],
  layout = c(6, 3)
) + create.scatterplot(
  fm2.predict ~ Days | Subject,
  data = sleepstudy,
  type = c('l'),
  lwd = 2.5,
  col = default.colours(3)[3],
  layout = c(6, 3))
```

## Sleep Study - Plot random effects
```{r, eval=FALSE, echo = T}
# Lattice dotplot
dotplot(ranef(fm2))
```

## Sleep Study - Plot random effects
```{r}
dotplot(ranef(fm2))
```

## Sleep Study - Compare Model 1 and 2
  - Use `anova(fm1, fm2)` for likelihood ratio test
```{r}
kableExtra::kable(broom::tidy(anova(fm1, fm2))[, -2], digits = c(0, 0, 0, 0, 0, 0, 0, 10))
```

## Sleep Study - Model 3 nlme
```{r, echo = T}
fm3.nlme <- lme(
  fixed = Reaction ~ 1 + Days,
  random = ~ 1 | Subject,
  data = sleepstudy,
  method = "ML")
```

## Sleep Study - Model 3 nlme
```{r, echo = T}
fm3.nlme
```

## Residual error structure
  - One benefit of using `nlme` is that we can specify error correlation structure
  - Using `lme4` we are constrain by the fact that $\boldsymbol{\epsilon} \sim N(0, \sigma I)$
  - With `nlme` we can model $\boldsymbol{\epsilon} \sim N(0, R)$, where $R$ can be specified.

## Sleep Study - Model 4 nlme with AR(1) errors
  - Assume that $\epsilon_t = \varphi \epsilon_{t - 1} + \delta_t$, where $\delta_t$ has zero mean and constant variance.
  - Including `correlation=corAR1(form = ~ 1 | Subject)` estimates $\varphi$.
  
```{r}
fm4.nlme <- lme(
  fixed = Reaction ~ 1 + Days,
  random = ~ 1 | Subject,
  correlation=corAR1(form = ~ 1 | Subject),
  data = sleepstudy,
  method = "ML")
```

## Sleep Study - Model 4 nlme with AR(1) errors
```{r}
fm4.nlme
```

## Sleep Study - Compare model 3 and 4
```{r}
kableExtra::kable(anova(fm3.nlme, fm4.nlme)[, -1], digits = c(0, 0, 0, 0, 0, 0, 1, 15))
```
